Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-11-10 MS_revised.tex
fixed table 2

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects \cite{mitter1988}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects \cite{nyman2012}. The
``escape-and-radiate'' hypothesis \cite{ehrlich1964}, the 
``oscillation hypothesis''
\cite{janz2011,nylin2014} or ``diffuse cospeciation'' \cite{nyman2012}.

@@

Although it has been suggested that part of the great diversity of
Nymphalidae butterflies is a result of hostplant-insect dynamics, it is
necessary to use modern techniques to investigate whether the
diversification patterns of Nymphalidae are in agreement with the
theoretical predictions. It is necessary to test whether the overall
diversification pattern of Nymphalidae is congruent with events of
sudden diversification bursts due to hostplant shift, climatic events or
shifts to closely related hostplants \cite{nylin2014,ferrer2013}.

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies \cite{wahlberg2009} to
investigate patterns
of diversification. We applied MEDUSA \cite{alfaro2009,harmon2011},
a recently developed
statistical method, to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA \cite{alfaro2009}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the method BiSSE as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}.

@@

@@

Patterns of diversification in Nymphalidae were analyzed by using MEDUSA
version 093.4.33 \cite{alfaro2009} on the maximum clade credibility tree from
Wahlberg et al. (2009) \cite{wahlberg2009}. MEDUSA fits alternative birth-death
likelihood models to a phylogenetic tree in order to estimate changes in
net diversification rates along branches. MEDUSA estimates likelihood
and AIC scores for the simplest birth-death model, with two parameters,
the rates \texttt{r}: net diversification \texttt{=} speciation
(\texttt{b}) \texttt{-} extinction (\texttt{d}) and \(\varepsilon\):
relative extinction \texttt{= d/b}. The AIC scores of the two-parameter
model are then compared with incrementally more complex models until the
addition of parameters do not improve the AIC scores beyond a cutoff
value. MEDUSA finds the likelihood of the models after taking into
account branch lengths and number of species per lineage 
\cite{alfaro2009}. Most
studies using MEDUSA only run the method on a single tree, usually the
maximum clade credibility tree, which makes the assumption that this
tree is correct. We wanted to study the effects of phylogenetic
uncertainty on estimation of net diversification rate shifts and
therefore selected 1000 random genus-level trees from the posterior
distribution of the Bayesian run from Wahlberg et al., 2009 \cite{wahlberg2009}
(MultiMEDUSA, supp. mat. 5). We calculated a new MCC tree from the
selection of trees, ran MEDUSA and MultiMEDUSA on the selection of 1000
trees and summarized the estimated changes in net diversification rates
for nodes across all trees. Patterns of change in net diversification
rates are considered significant if they are found at the same node in
at least 90\% of the trees. We also expected to find similar values of
net diversification (\texttt{r = b - d}) and relative extinction rate
($\varepsilon$ \texttt{= d/b}) values across the 1000 trees for the
nodes where changes in diversification tempo occurs. We used the AICc
threshold of 7.8 units, as estimated by MEDUSA, as the limit for a
significantly better fit to select among increasingly complex
alternative models.

@@

The MEDUSA analysis on the MCC tree in combination with richness data
estimated 18 changes in the tempo of diversification in Nymphalidae
history (Fig. 1; Table 1). The corrected AICc acceptance threshold for
adding subsequent piecewise birth-death processes to the overall model
was set to 7.8 units, as prescribed by MEDUSA. In all MEDUSA analyses,
the maximum number of inferred diversification splits in all trees was
26. The background net diversification rate for Nymphalidae was
estimated as \texttt{r = 0.092} lineages per Million of years and the
AICc score for the best fit model was \texttt{5090.5} (Table 1). MEDUSA
also estimated that the basic constant birth-death model was not a
better explanation for our data (AICc \texttt{= 5449.3}).

@@

For the diversification shifts found in both the MCC tree and most of
the samples of 1000 trees (frequency more than 90\%; Table 2), the
MultiMEDUSA approach recovered different rates of diversification than
those found using when the MCC tree alone.

There were four diversification shifts found with more than 0.90 of
probability in the trees from the random sample (Table 2): (i) net
diversification rate increase in the genus \emph{Ypthima}; (ii) net
diversification rate increase in the genus \emph{Charaxes}; (iii) rate
increase in Ithomiini subtribes Oleriina + Ithomiina + Napeogenina +
Dircennina + Godyridina; and (iv) rate increase in \emph{Callicore} +
\emph{Diaethria}.

MultiMEDUSA provided mean and standard deviation statistics for the
diversification values found on the shifts on the 1000 trees (supp. mat.
13--14), and found that some of the changes in net diversification rate
values had great variation across the posterior distribution of trees. A
boxplot of the net diversification rate values estimated for the clades
that appear in the MCC tree shows that some shifts are estimated as
increased or slowed diversification pace depending on the tree used for
analysis (Fig. 2). This variation is especially wide for the clade
formed by the genera \emph{Magneuptychia} and \emph{Caeruleuptychia}
because MEDUSA estimated diversification values from six times the
background net diversification rate (\texttt{r = 0.5234}) to almost zero
(\texttt{r = 1.22e-01}). The rates for \emph{Taenaris} were between 0.14
and 0.44 (mean value 0.25). Similar degrees of variation were found in
the nodes for \emph{Ypthima}, \emph{Charaxes} and \emph{Coenonympha}
(Fig. 2). The net diversification rates estimates for the clades
(Oleriina + Ithomiina + Napeogenina + Dircennina + Godyridina),
Limenitidinae + Heliconiinae and Pseudergolinae are relatively
consistent across the 1000 trees (Fig. 2).

@@

\section*{Conclusions}
We found that even though MEDUSA estimated several diversification
shifts in the maximum clade credibility tree of Nymphalidae, only a few
of these splits were found in more than 90\% of the trees from the
posterior distribution. In the literature, it is common practice that
conclusions are based on the splits estimated on the maximum clade
credibility tree. However, by using a MultiMEDUSA approach, we found
that for this Nymphalidae dataset some of these splits might be greatly
affected by phylogenetic uncertainty. Moreover, some of these splits can
be recovered either as increases or decreases in net diversification
rate depending on the tree from the posterior distribution that was used
for analysis. This means that contradictory conclusions would be made if
only the maximum clade credibility tree was used for analysis. We
recommend that all datasets should be analyzed using both approaches,
MEDUSA and MultiMEDUSA, in order to test whether the results are robust
when phylogenetic uncertainty is taken into account.

@@

 \item {\bf Supporting Information S08.pdf} Summary of the MulitMEDUSA
analysis on the 1000 trees from the posterior distribution.

@@

% Table 2
\begin{table}[!h]
    \caption{\bf{Differences in rates estimated by MEDUSA on the MCC
tree from Wahlberg and the MultiMEDUSA approach on 1000 random trees
from the posterior distribution.}}
\begin{tabular}{lcrr}
Split.Node & rate by MEDUSA & Median rate by MultiMEDUSA & probability of being recovered \\
1          & 0.090          & not found                  & 0.000                          \\
2          & 0.054          & -0.030                     & 0.864                          \\
3          & 0.311          & 0.219                      & 0.961                          \\
4          & 0.290          & 0.212                      & 0.996                          \\
5          & 0.187          & 0.101                      & 0.927                          \\
6          & 0.116          & 0.039                      & 0.131                          \\
7          & 0.064          & -0.052                     & 0.195                          \\
8          & 0.240          & 0.166                      & 0.319                          \\
9          & 0.042          & -0.049                     & 0.897                          \\
10         & 0.209          & 0.149                      & 0.619                          \\
11         & 0.311          & 0.208                      & 0.831                          \\
12         & 0.219          & 0.135                      & 0.911                          \\
13         & 0.220          & 0.117                      & 0.276                          \\
14         & 0.190          & 0.115                      & 0.379                          \\
15         & 0.234          & 0.127                      & 0.825                          \\
16         & 0.000          & -0.005                     & 0.659                          \\
17         & 0.311          & 0.248                      & 0.785                          \\
18         & 0.024          & -0.064                     & 0.024                          \\
19         & 0.000          & -0.087                     & 0.381                         
\end{tabular}
\end{table}

>>>>

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects \cite{mitter1988}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects \cite{nyman2012}. The
``escape-and-radiate'' hypothesis \cite{ehrlich1964}, the 
``oscillation hypothesis''
\cite{janz2011,nylin2014} or ``diffuse cospeciation'' \cite{nyman2012} and the
``musical chairs hypothesis'' \cite{hardy2014}.

@@

Although it has been suggested that part of the great diversity of
Nymphalidae butterflies is a result of hostplant-insect dynamics, it is
necessary to use modern techniques to investigate whether the
diversification patterns of Nymphalidae are in agreement with the
theoretical predictions. It is necessary to test whether the overall
diversification pattern of Nymphalidae is congruent with events of
sudden diversification bursts due to hostplant shift or climatic events
\cite{nylin2014,ferrer2013}.

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies \cite{wahlberg2009} to
investigate patterns
of diversification. We applied the statistical method
MEDUSA \cite{alfaro2009,harmon2011},
to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA \cite{alfaro2009}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the method BiSSE as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}.

@@

We analyzed our MCC tree in the Bayesian software BAMM in order to contrast
the results
found by MEDUSA and MultiMEDUSA.
BAMM (Bayesian analysis of macroevolutionary mixtures)
\cite{rabosky2014} is a
software that fits several models of speciation and extinction. The best model
can be selected by comparison of Bayes Factors.

@@

Patterns of diversification in Nymphalidae were analyzed by using MEDUSA
version 093.4.33 \cite{alfaro2009} on the maximum clade credibility tree from
Wahlberg et al. (2009) \cite{wahlberg2009}. MEDUSA fits alternative birth-death
likelihood models to a phylogenetic tree in order to estimate changes in
net diversification rates along branches. MEDUSA estimates likelihood
and AIC scores for the simplest birth-death model, with two parameters,
the rates \texttt{r}: net diversification \texttt{=} speciation
(\texttt{b}) \texttt{-} extinction (\texttt{d}) and ($\varepsilon$:
relative extinction \texttt{= d/b}. The AIC scores of the two-parameter
model are then compared with incrementally more complex models until the
addition of parameters do not improve the AIC scores beyond a cutoff
value. MEDUSA finds the likelihood of the models after taking into
account branch lengths and number of species per lineage 
\cite{alfaro2009}. Most
studies using MEDUSA only run the method on a single tree, usually the
maximum clade credibility tree, which makes the assumption that this
tree is correct. We wanted to study the effects of phylogenetic
uncertainty on estimation of net diversification rate shifts and
therefore selected 1000 random genus-level trees from the posterior
distribution of the Bayesian run from Wahlberg et al., 2009 \cite{wahlberg2009}
(MultiMEDUSA, supp. mat. 5). We calculated a new MCC tree from the
selection of trees, ran MEDUSA and MultiMEDUSA on the selection of 1000
trees and summarized the estimated changes in net diversification rates
for nodes across all trees. Patterns of change in net diversification
rates are considered significant if they are found at the same node in
at least 90\% of the trees. We also expected to find similar values of
net diversification (\texttt{r = b - d}) and relative extinction rate
($\varepsilon$ \texttt{= d/b}) values across the 1000 trees for the
nodes where changes in diversification tempo occurs. We used the AICc
threshold of 7.8 units, as estimated by MEDUSA, as the limit for a
significantly better fit to select among increasingly complex
alternative models.

@@

The MEDUSA analysis on the MCC tree in combination with richness data
estimated 18 significant changes in the tempo of diversification in Nymphalidae
history (Fig. 1; Table 1). The corrected AICc acceptance threshold for
adding subsequent piecewise birth-death processes to the overall model
was set to 7.8 units, as prescribed by MEDUSA. In all MEDUSA analyses,
the maximum number of inferred diversification splits in all trees was
26. The background net diversification rate for Nymphalidae was
estimated as \texttt{r = 0.092} lineages per Million of years and the
AICc score for the best fit model was \texttt{5090.5} (Table 1). MEDUSA
also estimated that the basic constant birth-death model was not a
better explanation for our data (AICc \texttt{= 5449.3}).

@@

For the diversification shifts found in both the MCC tree and most of
the samples of 1000 trees (frequency more than 90\%; Table 2), the
MultiMEDUSA approach recovered different rates of diversification than
those found when the MCC tree alone.

There were four diversification shifts found in the trees from the
random sample (Table 2): (i) net
diversification rate increase in the genus \emph{Ypthima} ($r = 0.22$);
(ii) net diversification rate increase in the genus \emph{Charaxes}
($r = 0.21$); (iii) rate increase in Ithomiini subtribes Oleriina +
Ithomiina + Napeogenina + Dircennina + Godyridina ($r = 0.10$);
and (iv) rate increase in \emph{Callicore} + \emph{Diaethria} ($r = 0.135$).

MultiMEDUSA provided mean and standard deviation statistics for the
diversification values on the shifts on the 1000 trees (supp. mat.
13--14), and found that some of the changes in net diversification rate
values had great variation across the posterior distribution of trees. A
boxplot of the net diversification rate values estimated for the clades
that appear in the MCC tree shows that some shifts are estimated as
increased or slowed diversification pace depending on the tree used for
analysis (Fig. 2). This variation is especially wide for the clade
formed by the genera \emph{Magneuptychia} and \emph{Caeruleuptychia}
because MEDUSA estimated diversification values from six times the
background net diversification rate (\texttt{r = 0.5234}) to almost zero
(\texttt{r = 1.22e-01}). The rates for \emph{Taenaris} were between 0.14
and 0.44 (mean value 0.25). Similar degrees of variation were found in
the nodes for \emph{Ypthima}, \emph{Charaxes} and \emph{Coenonympha}
(Fig. 2). The net diversification rates estimates for the clades
(Oleriina + Ithomiina + Napeogenina + Dircennina + Godyridina),
Limenitidinae + Heliconiinae and Pseudergolinae are relatively
consistent across the 1000 trees (Fig. 2).

@@

\section*{Conclusions}
We found that even though MEDUSA estimated several diversification
shifts in the maximum clade credibility tree of Nymphalidae, only a few
of these splits were found in more than 90\% of the trees from the
posterior distribution. In the literature, it is common practice that
conclusions are based on the splits estimated on the maximum clade
credibility tree. However, by using a MultiMEDUSA approach, we found
that for this Nymphalidae dataset some of these splits might be greatly
affected by phylogenetic uncertainty. Moreover, some of these splits can
be recovered either as increases or decreases in net diversification
rate depending on the tree from the posterior distribution that was used
for analysis. This means that contradictory conclusions would be made if
only the maximum clade credibility tree was used for analysis. We
recommend that all datasets should be analyzed using both approaches,
MEDUSA and MultiMEDUSA, in order to test whether the results are robust
when phylogenetic uncertainty is taken into account.
Moreover, we also found that the results from BAMM, without taking into account
phylogenetic uncertainty, correspond better to those of MultiMEDUSA than MEDUSA.

@@

 \item {\bf Supporting Information S08.pdf} Summary of the MultiMEDUSA
analysis on the 1000 trees from the posterior distribution.

@@

% Table 2
\begin{table}[!h]
    \caption{\bf{Differences in rates estimated by MEDUSA on the MCC
tree from the sample of trees from the posterior distribution and the
MultiMEDUSA approach. Shift consistently recovered across the
sample of trees in bold face.}}
\begin{tabular}{lcrr}
Split.Node & rate by MEDUSA & Median rate by MultiMEDUSA & probability of being recovered \\
1          & 0.092          & not found                  & 0.000                          \\
2          & 0.055          & -0.030                     & 0.864                          \\
\bf{3}     & 0.184          & \bf{0.219}                 & \bf{0.961}                     \\
\bf{4}     & 0.166          & \bf{0.212}                 & \bf{0.996}                     \\
\bf{5}     & 0.111          & \bf{0.101}                 & \bf{0.927}                     \\
6          & 0.119          & 0.039                      & 0.131                          \\
7          & 0.066          & -0.052                     & 0.195                          \\
8          & 0.232          & 0.166                      & 0.319                          \\
9          & 0.042          & -0.049                     & 0.897                          \\
10         & 0.058          & 0.149                      & 0.619                          \\
11         & 0.311          & 0.208                      & 0.831                          \\
\bf{12}    & 0.219          & \bf{0.135}                 & \bf{0.911}                     \\
13         & 0.082          & 0.117                      & 0.276                          \\
14         & 0.099          & 0.115                      & 0.379                          \\
15         & 0.113          & 0.127                      & 0.825                          \\
16         & 0.222          & -0.005                     & 0.659                          \\
17         & 0.243          & 0.248                      & 0.785                          \\
18         & 0.192          & -0.064                     & 0.024                          \\
19         & 0.064          & -0.087                     & 0.381                         
\end{tabular}
\end{table}
